Citation Analysis of Computer Science Publications

### 2019-12-10 Scientometrics/comp.tex
added Perianes-Rodriguez

\begin{abstract}
Computer science as a field has experienced dramatic growth and diversification over the last twenty years. Towards an improved understanding of the structure of the field, we analyze a cohort of the computer science literature.  For insight on the features of this cohort and the flow of information within its components, we construct article level clusters whose elements are linked by citations or co-citations, and reconcile them with major and minor subject categories in the All Science Journal Classification (Scopus). [abstract is feeble and needs to be rewritten]

@@

Computer science, and its applications, has experienced rapid growth and diversification over the last twenty years. More recently, the collective influence of the Internet of Things (IoT), `big' data, accessible cloud computing, and advances in artificial intelligence have been postulated as a recent driver for growth and evolution~\cite{siebel2019_digital}. As noted in a 2017 US National Academies Report~\cite{nas_2017}, ``A wide range of jobs in virtually all sectors demand computing skills to an unprecedented extent. And every academic discipline finds itself incorporating computing into its research and educational mission". Given this powerful influence some understanding of the present state and structure of computer science and its relationship to other fields would inform planning and policy making at multiple levels from national level funding all the way down to faculty hiring strategy. For the same reason of broad influence, computer science can also serve as a model system for studying interdisciplinarity.

Noting that that the scientific literature serves a rich source of information to study the structure and historical development of a field, Salton and Bergmark conducted a historical study  of the computer science literature (419 computer articles published in 1974,  and 3,812 references cited in these articles)~\cite{salton_citation_1979}. These authors described the global structure of computer science as comprising three main subareas (i) theoretical foundations, such as theory of computation, (ii) hardware and computer systems, such as architecture,  and (iii) software, such as programming systems.  Related areas noted were (a) mathematics of computing, such as numerical analysis, (b) special software topics, such as operating systems, (c) data management and database systems, (d) methodologies valid for multiple applications, such as algebraic manipulation, (e) computer applications, such as computer graphics, (f) non-technical aspects, such as computer education.

Computing has grown and evolved significantly since 1974. Progressing beyond Salton and Bergmark's historical triad of theoretical foundations; hardware and computer systems; and software, the Computing Classification System (CCS)~\cite{acm_ref} published by the Association for Computing Machinery now consists of 13 top-level areas that reflect a more current view of the field. This classification also addresses relationships with other fields under the category of Applied Computing. However, an easy way to map scientific publications, especially interdisciplinary articles or those from adjacent fields, to the CCS classification does not exist.

Other classification systems are available such as the All Science Journal Classification (ASJC) developed and maintained by Scopus, the Web of Science research and categorical classifications from Clarivate Analytics, and the NSF classification system~\cite{nsf_classification}. Of these, Scopus and the Web of Science also include global bibliographies with citation links and proprietary unique identifiers for publications, some of which can be matched to digital object identifiers (DOIs). All three rely on applying one or more journal-derived labels to articles. The inheritance, by articles, of labels meant to describe journals can be problematic in fields that evolve faster than their journal classifications. Shu and colleagues recently noted in a comparative study of the Chinese Science Citation Database (CSCD) and the Web of Science, that 46\% of articles did not belong to the discipline of the journal they were published in~\cite{shu_comparing_2019}. Others have also discussed and critiqued disciplinary assignments using journal based classifications~\cite{wang_large-scale_2016} \textcolor{blue}{need comprehensive list of citations}. Article classification systems have been constructed, however, which escape some of the criticisms of journal-based classification~\cite{traag_louvain_2019,boyack_classification_2014,waltman_new_2012}.

In this study, we sought to extend the work of Salton and Bergmark to reflect a more current reality while also considering connections between computer science and other fields. As a source of computer science literature, we used DBLP, a reference bibliography for computer science~\cite{dblp_ref}. The DBLP bibliography covers publications from computer science and includes publications from hybrid fields, where they are considered pertinent to computer science research. The underlying assumption is that records in DBLP are greatly enriched for computer science even while recognizing that coverage may be incomplete and possibly biased. Cross-matching DBLP publications to Scopus records enabled us to harvest the richer citation data in Scopus, as well as extract links to publications from other disciplines. This cross-matching also allows the use of both journal and article based classifications when clustering documents. Using both direct citations and co-citations as the basis for clustering, we used DBLP articles from journals and conference proceedings for the period 1996-2015. We reconciled these clusters with the All Science Journal Classification (ASJC) developed and maintained by Scopus through a combination of automated and manual procedures. We describe below the high-level interactions of computer science within the Physical Sciences as well as with the Social Sciences and Life Sciences.

@@

\paragraph{Clustering by Direct Citation.} Graclus is a spectral graph clustering package that computes normalized cut and ratio association for a given undirected graph~\cite{graclus_2007}.  We used v1.2 in our experiments. The \emph{comp} dataset was formatted as an undirected graph, stored in a file with a header line indicating the number of nodes and edges, and used as input to Graclus, which requires the number of clusters to be formed as an input parameter. In preliminary experiments, we varied the number of clusters to be formed between 10 and 50 clusters (data not shown).  At 20 clusters, the most evenly sized set of clusters resulted with the largest cluster containing roughly 10 times the number of nodes in the smallest one~\cite{traag_louvain_2019}.  Consequently, we generated 20 clusters, labeled 0-19, using Graclus (Fig~\ref{fig:heatmap}).
We used conductance, as defined in Shun et al.~\cite{shun_parallel_2016}, to  evaluate clustering (smaller is better), noting that
conductance has been found to be ``the stand-alone quality metric that best indicates performance on the information recovery metrics" \cite{emmons2016analysis}.
% The quality of clustering was evaluated by measuring conductance as defined in Shun et al.~\cite{shun_parallel_2016}. 
%George, Tandy - add text about why conductance is good, and that small values are better
The data in \emph{comp} were clustered into 18, 20, or 22 clusters and in each case the highest numbered cluster (17, 19, or 21) had the greatest conductance value and the smallest number of nodes, suggesting that it effectively served as a container for `left over publications' during the Graclus process. Thus, we did not consider cluster 19 (the 20th cluster) when interpreting our results.  The remaining 19 clusters had conductance values ranging from 0.09 to 0.25 with a median conductance of 0.15.

 \begin{figure}[ht]
\centering
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.5]{graclus_comparison.pdf}
% figure caption is below the figure
\caption{Summary of DBLP data cross-matched with Scopus. 2,685,356 publications from DBLP were cross-matched with Scopus and then grouped by the 27 major subject areas in the ASJC (Scopus) classification. The largest number of publications are contributed by Computer Science; Mathematics; and Engineering and then by Social Sciences; Physics and Astronomy; and Biochemistry, Genetics, and Molecular Biology. Publications were further annotated with respect to being either articles \emph{(ar)} or conference proceedings \emph{(cp)}. For this dataset, the major subject areas of Computer Science; Engineering; and Mathematics contributed the most publications while Pharmacology, Toxicology, and Pharmaceutics; Nursing, Veterinary and Dentistry contributed the least.}
\label{fig:graclus_comparison}       % Give a unique label
\end{figure}

\begin{table}
% table caption is above the table
\caption{\emph{Statistics regarding the 20 clusters produced using direct citation.
2685021 is because of inner joins on ASJC Explain conductance calculations~\cite{shun_parallel_2016}}}
\label{tab:1}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{lrccc}
\hline\noalign{\smallskip}
Cluster & Publications & Conductance & Total ASJC Labels & Unique ASJC Labels\\
\noalign{\smallskip}\hline\noalign{\smallskip}
0 & 111,294 & 0.12 & 265,664 & 142 \\ 
1 & 117,057 & 0.14 & 246,960 & 166 \\ 
2 & 116,251 & 0.11 & 280,602 & 165 \\ 
3 & 353,366 & 0.15 & 881,693 & 200 \\ 
4 & 145,081 & 0.09 & 349,020 & 154 \\ 
5 & 92,097 & 0.17 & 248,168 & 186 \\ 
6 & 71,865 & 0.15 & 199,681 & 163 \\ 
7 & 179,927 & 0.18 & 465,181 & 202 \\ 
8 & 302,656 & 0.18 & 760,117 & 214 \\ 
9 & 69,520 & 0.19 & 197,031 & 174 \\ 
10 & 42,462 & 0.10 & 102,838 & 141 \\ 
11 & 448,030 & 0.25 & 1,229,061 & 224 \\ 
12 & 70,738 & 0.21 & 216,546 & 179 \\ 
13 & 105,232 & 0.17 & 289,318 & 187 \\ 
14 & 199,176 & 0.15 & 551,657 & 208 \\ 
15 & 64,384 & 0.17 & 195,679 & 167 \\ 
16 & 89,340 & 0.11 & 240,157 & 158 \\ 
17 & 50,817 & 0.14 & 181,531 & 179 \\ 
18 & 43,113 & 0.20 & 108,518 & 177 \\ 
19 & 12,615 & 0.80 & 36,583 & 229 \\ \noalign{\smallskip}\hline
\end{tabular}
\end{table}

>>>>

\begin{abstract}
Computer science as a field has experienced dramatic growth and diversification over the last twenty years. Towards an current understanding of the structure of the field, we analyze a cohort of the computer science literature.  For insight on the features of this cohort and the flow of information within its components, we construct article level clusters based on either direct citations or co-citations, and reconcile them with major and minor subject categories in the All Science Journal Classification (Scopus). [abstract is feeble and needs to be rewritten].

@@

Computer science, and its applications, has experienced rapid growth and diversification over the last twenty years. As observed in a 2017 US National Academies Report, ``A wide range of jobs in virtually all sectors demand computing skills to an unprecedented extent. And every academic discipline finds itself incorporating computing into its research and educational mission"~\cite{nas_2017}. More recently, the collective influence of the Internet of Things (IoT), `big' data, accessible cloud computing, and advances in artificial intelligence have been postulated as a recent driver for growth and evolution~\cite{siebel2019_digital}. Given this powerful influence some understanding of the present state and structure of computer science and its relationship to other fields could inform planning and policy making at multiple levels from national level funding all the way down to faculty hiring strategy. For the same reason of broad influence, computer science can also serve as a model system for studying interdisciplinarity.

Noting that that the scientific literature serves a rich source of information to study the structure and historical development of a field, Salton and Bergmark conducted a historical study  of the computer science literature (419 computer articles published in 1974,  and 3,812 references cited in these articles)~\cite{salton_citation_1979}. These authors described the global structure of computer science as comprising three main areas (i) theoretical foundations, such as theory of computation, (ii) hardware and computer systems, such as architecture,  and (iii) software, such as programming systems.  Related areas noted were (a) mathematics of computing, such as numerical analysis, (b) special software topics, such as operating systems, (c) data management and database systems, (d) methodologies valid for multiple applications, such as algebraic manipulation, (e) computer applications, such as computer graphics, (f) non-technical aspects, such as computer education.

Progressing beyond 1974-79, and Salton and Bergmark's historical triad of theoretical foundations; hardware and computer systems; and software, the Computing Classification System (CCS)~\cite{acm_ref} published by the Association for Computing Machinery now consists of 13 top-level areas that reflect a more current view of the field. This classification also addresses relationships with other fields under the category of Applied Computing. However, an easy way to map scientific publications, especially interdisciplinary articles or those from fields proximal to the CCS classification does not exist.

Other classification systems are available such as the All Science Journal Classification (ASJC) developed and maintained by Scopus, the Web of Science research and categorical classifications from Clarivate Analytics, and the NSF classification system~\cite{nsf_classification}. All three rely on applying one or more journal-derived labels to articles.  Of these, Scopus and the Web of Science also include global bibliographies with citation links and proprietary unique identifiers for publications, some of which can be matched to digital object identifiers (DOIs). Applying labels, designed for journals, to articles, of labels meant can be problematic in uterms of accuracy and because fields evolve faster than their journal classifications. [should we put in a comment about why article level labels provided by authors have limited use and that enforcement of controlled vocabularies varies by publisher?] Shu and colleagues recently noted in a comparative study of the Chinese Science Citation Database (CSCD) and the Web of Science, that 46\% of articles did not belong to the discipline of the journal they were published in~\cite{shu_comparing_2019}. Others have also discussed and critiqued disciplinary assignments using journal based classifications~\cite{wang_large-scale_2016,perianes-rodriguez_comparison_2017} \textcolor{blue}{need comprehensive list of citations}. Article classification systems have been constructed, however, which escape some of the criticisms of journal-based classification~\cite{traag_louvain_2019,boyack_classification_2014,waltman_new_2012} but seem to be in the research domain rather than the domain of practical use [need to check on this].[May need a nother bridging sentence to the next paragraph].

In this study, we sought to extend the work of Salton and Bergmark to reflect a more current reality while also considering connections between computer science and other fields. As a source of computer science literature, we used DBLP, a reference bibliography for computer science~\cite{dblp_ref}. The DBLP bibliography covers publications from computer science and includes publications from hybrid fields, where they are considered pertinent to computer science research. The underlying assumption we made is that records in DBLP are greatly enriched for computer science even while recognizing that coverage may be incomplete and possibly biased. Cross-matching DBLP publications to records in the Scopus, abstract and citation database of peer-reviewed literature, enabled us to harvest the richer links in Scopus as well as extract links to publications from other disciplines. This cross-matching also allows the use of both journal and article based classifications when clustering documents. Using both direct citations and co-citations as the basis for clustering, we used DBLP articles from journals and conference proceedings for the period 1996-2015. We reconciled these clusters with the All Science Journal Classification (ASJC) developed and maintained by Scopus through a combination of automated and manual procedures. We describe below the high-level interactions of computer science within the Physical Sciences as well as with the Social Sciences and Life Sciences. [What did we find? Some kind of preview sentence!].

@@

\paragraph{Clustering by Direct Citation.} Graclus is a spectral graph clustering package that computes normalized cut and ratio association for a given undirected graph~\cite{graclus_2007}.  We used v1.2 in our experiments. The \emph{comp} dataset was formatted as an undirected graph, stored in a file with a header line indicating the number of nodes and edges, and used as input to Graclus, which requires the number of clusters to be formed as an input parameter. In preliminary experiments, we varied the number of clusters to be formed between 10 and 50 clusters (data not shown).  At 20 clusters, the most evenly sized set of clusters resulted with the largest cluster containing roughly 10 times the number of nodes in the smallest one~\cite{traag_louvain_2019}.  Consequently, we generated 20 clusters, labeled 0-19, using Graclus (Fig~\ref{fig:heatmap}).
We used conductance, as defined in Shun et al.~\cite{shun_parallel_2016}, to  evaluate clustering (smaller is better), noting that
conductance has been found to be ``the stand-alone quality metric that best indicates performance on the information recovery metrics" \cite{emmons2016analysis}. Data from \emph{comp} were clustered into 18, 20, or 22 clusters and in each case the highest numbered cluster (17, 19, or 21) had the greatest conductance value and the smallest number of nodes, suggesting that it effectively served as a container for `left over publications' during the Graclus process. Thus, we did not consider cluster 19 (the 20th cluster) when interpreting our results.  The remaining 19 clusters had conductance values ranging from 0.09 to 0.25 with a median conductance of 0.15. [This section on comp and conductneeds to be reasoned and presented better- for example why do we think these are good conductance values?].

\begin{figure}[ht]
\centering
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.5]{graclus_comparison.pdf}
% figure caption is below the figure
\caption{Summary of DBLP data cross-matched with Scopus. 2,685,356 publications from DBLP were cross-matched with Scopus and then grouped by the 27 major subject areas in the ASJC (Scopus) classification. The largest number of publications are contributed by Computer Science; Mathematics; and Engineering and then by Social Sciences; Physics and Astronomy; and Biochemistry, Genetics, and Molecular Biology. Publications were further annotated with respect to being either articles \emph{(ar)} or conference proceedings \emph{(cp)}. For this dataset, the major subject areas of Computer Science; Engineering; and Mathematics contributed the most publications while Pharmacology, Toxicology, and Pharmaceutics; Nursing, Veterinary and Dentistry contributed the least.}
\label{fig:graclus_comparison}       
\end{figure}

\begin{table}
\caption{\emph{Statistics regarding the 20 clusters produced using direct citation. The \emph{comp} dataset 
2685021 is because of inner joins on ASJC Explain conductance calculations~\cite{shun_parallel_2016}}}
\label{tab:1}       
\begin{tabular}{lrccc}
\hline\noalign{\smallskip}
Cluster & Publications & Conductance & Total ASJC Labels & Unique ASJC Labels\\
\noalign{\smallskip}\hline\noalign{\smallskip}
0 & 111,294 & 0.12 & 265,664 & 142 \\ 
1 & 117,057 & 0.14 & 246,960 & 166 \\ 
2 & 116,251 & 0.11 & 280,602 & 165 \\ 
3 & 353,366 & 0.15 & 881,693 & 200 \\ 
4 & 145,081 & 0.09 & 349,020 & 154 \\ 
5 & 92,097 & 0.17 & 248,168 & 186 \\ 
6 & 71,865 & 0.15 & 199,681 & 163 \\ 
7 & 179,927 & 0.18 & 465,181 & 202 \\ 
8 & 302,656 & 0.18 & 760,117 & 214 \\ 
9 & 69,520 & 0.19 & 197,031 & 174 \\ 
10 & 42,462 & 0.10 & 102,838 & 141 \\ 
11 & 448,030 & 0.25 & 1,229,061 & 224 \\ 
12 & 70,738 & 0.21 & 216,546 & 179 \\ 
13 & 105,232 & 0.17 & 289,318 & 187 \\ 
14 & 199,176 & 0.15 & 551,657 & 208 \\ 
15 & 64,384 & 0.17 & 195,679 & 167 \\ 
16 & 89,340 & 0.11 & 240,157 & 158 \\ 
17 & 50,817 & 0.14 & 181,531 & 179 \\ 
18 & 43,113 & 0.20 & 108,518 & 177 \\ 
19 & 12,615 & 0.80 & 36,583 & 229 \\ \noalign{\smallskip}\hline
\end{tabular}
\end{table}
